Comparison of dimensionality reduction schemes for parallel global optimization algorithms \thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150}

### 2018-04-08 2018/RSD/MultiEvolvents/RSD.tex
RSD: update numerical experiments

\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbf{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbf{R}^N\) space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .
\par
As it has been shown earlier by many researchers
(see, for instance, \cite{floudasPardalosGOState}, \cite{horstTuyGO}, \cite{pinterGO}, \cite{strSergGO}),
finding the numerical estimate of the global optimum implies constructing a coverage of
the search domain \(D\). As a result, the computational costs of solving the global
optimization problems are readily very high even for a small number of varied parameters
(the dimensionality of the problem). A notable reduction in the volume of computations
can be achieved when the coverage of the search domain is non-uniform, i. e. the series
of trial points is only dense in a vicinity of the global optimum point. The construction
of such a non-uniform coverage could be provided in an adaptive way, when the selection
of the next trial points is determined by using the search information (the preceding
trial points and the values of the minimized function at these points) obtained in the course
of computations. This necessary condition complicates considerably the computational schemes
of global optimization methods since it implies a complex analysis of a large amount
of multidimensional search information. As a result, many optimization algorithms use
various approaches to the dimensional reduction \cite{pinterGO}, \cite{sergeyevStronginLera2013}, \cite{strongin1978}, \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}.

@@

\section{Results of Numerical Experiments}
The computational experiments have been carried out on the Lobachevsky supercomputer at
State University of Nizhny Novgorod. A computational node included 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. The CPUs had 8 cores (i. e. total 16 cores
were available per a node). Все рассматриваемые алгортимы и развёртки были реализованы на языке C++ в рамках программной системы Globalizer.
Для реализации параллелизма на одном узле использована технология OpenMP, а для параллелизма на несколько узлов --- стандарт MPI.

@@

\begin{table}
\begin{center}
\caption{Среднее количество вычислений \(g_0\) и \(\varphi\) при решении задач класса GKLS 3d Simple с помощью сдвиговой развёртки}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
  $L$ & Constraint & Objective (shifted) & $\frac{calc(g_0)}{calc(\varphi)}$ ratio \\
  \hline
  2 & 96247.9  & 6840.14 & 14.07\\
  \hline
  3 & 153131.0 & 7702.82 & 19.88\\
  \hline
  \end{tabular}
  \label{tab:shifted_g0}
\end{center}
\end{table}

\subsection{Параллельные вращаемые развертки}
\begin{table}
  \centering
  \caption{Averaged numbers of iterations executed by GPMAGS for solving the test optimization problems}
  \label{tab:4}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
    \noalign{\smallskip}\hline
    I &
    \parbox{0.25\textwidth}{
    \begin{center}
    \textbf{1 cluster node}
    \end{center}		}
      & 1 & 11953 & 25263 & & 15920 & \(>\)148342 (4)  \\
    &  & 32 & 11953 & 25263 & & 15920 & \(>\)148342 (4)  \\
    \hline \noalign{\smallskip}
II  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & 1 & 4762 & 11178 & & 13378 & 109075 \\
&   & 32 & 2372 & 5972 & & 5203 & 51868 \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & 1  & 171 & 393 & & 382 & 3452  \\
&    & 32 & 85 & 182 & & 249 & 1306 \\

    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Speedup of parallel computations executed by GMAGS}
  \label{tab:4}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
    \noalign{\smallskip}\hline
    I &
    \parbox{0.25\textwidth}{
    \begin{center}
    \textbf{1 cluster node}
    \end{center}		}
      & 1 & 11953 & 25263 & & 15920 & \(>\)148342 (4)  \\
    &  & 32 & 11953 & 25263 & & 15920 & \(>\)148342 (4)  \\
    \hline \noalign{\smallskip}
II  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & 1 & 4762 & 11178 & & 13378 & 109075 \\
&   & 32 & 2372 & 5972 & & 5203 & 51868 \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & 1  & 171 & 393 & & 382 & 3452  \\
&    & 32 & 85 & 182 & & 249 & 1306 \\

    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

\section{Вывод о целесообразности применения того или иного вида разверток для того или иного вида задач}
s

@@

>>>>

\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer\cite{globalizerSystem}, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbf{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbf{R}^N\) space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .
\par
As it has been shown earlier by many researchers
(see, for instance, \cite{floudasPardalosGOState}, \cite{horstTuyGO}, \cite{pinterGO}, \cite{strSergGO}),
finding the numerical estimate of the global optimum implies constructing a coverage of
the search domain \(D\). As a result, the computational costs of solving the global
optimization problems are readily very high even for a small number of varied parameters
(the dimensionality of the problem). A notable reduction in the volume of computations
can be achieved when the coverage of the search domain is non-uniform, i. e. the series
of trial points is only dense in a vicinity of the global optimum point. The construction
of such a non-uniform coverage could be provided in an adaptive way, when the selection
of the next trial points is determined by using the search information (the preceding
trial points and the values of the minimized function at these points) obtained in the course
of computations. This necessary condition complicates considerably the computational schemes
of global optimization methods since it implies a complex analysis of a large amount
of multidimensional search information. As a result, many optimization algorithms use
various approaches to the dimensional reduction \cite{pinterGO}, \cite{sergeyevStronginLera2013}, \cite{strongin1978}, \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}.

@@

\section{Results of Numerical Experiments}
The computational experiments have been carried out on the Lobachevsky supercomputer at
State University of Nizhny Novgorod. A computational node included 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. The CPUs had 8 cores (i. e. total 16 cores
were available per a node). Все рассматриваемые алгортимы и развёртки были реализованы на языке C++ в рамках программной системы Globalizer\cite{globalizerSystem}.
Для реализации параллелизма на одном узле использована технология OpenMP, а для параллелизма на несколько узлов --- стандарт MPI.

@@

\begin{table}
\begin{center}
\caption{Среднее количество вычислений \(g_0\) и \(\varphi\) при решении задач класса GKLS 3d Simple с помощью сдвиговой развёртки}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
  $L$ & $calc(g_0)$ & $calc(\varphi)$ & $\frac{calc(g_0)}{calc(\varphi)}$ ratio \\
  \hline
  2 & 96247.9  & 6840.14 & 14.07\\
  \hline
  3 & 153131.0 & 7702.82 & 19.88\\
  \hline
  \end{tabular}
  \label{tab:shifted_g0}
\end{center}
\end{table}

\subsection{Параллельные вращаемые развертки}
Чтобы оценить эффективность параллельного алгортима из секции \ref{sec:parallel_evolvents} были проведены вычислительные эксперименты на классах GKLS 4d (Hard, Simple) и GKLS 5d (Hard, Simple). Значения \(r\) во всех экспериментах равно 5.0, размер \(\delta\)-окрестности известного решения увеличен до 0.3. При решении серий задач использовалось до 8 узлов кластера и до 32 вычислительных потоков на каждом узле.

В таблице \ref{tab:iterations} приведено среднее количество итераций при решении 100 задач из каждого из рассматриваемых классов. При увеличении числа узлов и числа потоков на каждом узле количество итераций заметно сокращается (за исключением класса GKLS 4d Simple при переходе с 1 узла на 4 узла в однопоточном режиме).

\begin{table}
  \centering
  \caption{Averaged numbers of iterations executed by the parallel algorithm for solving the test optimization problems}
  \label{tab:iterations}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
    \noalign{\smallskip}\hline
    I &
    \parbox{0.25\textwidth}{
    \begin{center}
    \textbf{1 cluster node}
    \end{center}		}
      & \textit{1} & 12167 & 25635 & & 20979 & 187353  \\
    &  & \textit{32} & 328 & 1268  & &   898 & 12208 \\
    \hline \noalign{\smallskip}
II  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1} & 25312 & 11103 & & 1472 & 17009 \\
&   & \textit{32} & 64 &   913 & & 47 & 345 \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 810 & 4351 & & 868 & 5697  \\
& & \textit{32} & 34  & 112  & & 35  & 868 \\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

Если считать, что затраты на параллелизм пренебрежимо малы по сравнению с затратами на вычисление целевых функций в задачах оптимизации, то ускорение по времени от использования параллельного метода будет равно ускорению по итерациям. Однако, в действительности это предположение не всегда справедливо. Во всех численных экспериментах время вычисления целевой функции занимает примерно $10^{-3}$с. В таблице \ref{tab:speedup} приведено ускорение по итерациям и в куглых скобках ускорение по времени. В первой строке таблицы, соответствующей последовательному режиму, в скобках приведено среднее время решения одной задачи. Из таблицы видно, что для класов GKLS 4d выгоднее использовать один узел в многопоточном режиме, тогда как для решения более сложных пятимерных задач лучше использовать несколько узлов, каждый из которых работает в параллельном режиме.

\begin{table}
  \centering
  \caption{Speedup of parallel computations executed by the parallel algorithm}
  \label{tab:speedup}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
    \noalign{\smallskip}\hline
    I &
    \parbox{0.25\textwidth}{
    \begin{center}
    \textbf{1 cluster node}
    \end{center}		}
    & \textit{1}   & 12167(10.58s) & 25635(22.26s) & & 20979(22.78s) & 187353(205.83s)  \\
  &  & \textit{32} & 37.1(18.03) & 20.2(8.55)  & &  23.3(8.77) & 15.4(9.68) \\
  \hline \noalign{\smallskip}
II  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
& \textit{1} &        0.5(0.33) & 2.3(0.86)  & & 14.3(6.61) & 11.0(6.06) \\
&   & \textit{32} & 190.1(9.59) & 28.1(1.08) & & 446.4(19.79) & 543.0(43.60) \\
  \noalign{\smallskip}\hline	\noalign{\smallskip}
III & \textbf{8 cluster nodes} %\multirow{3}{*}{}
& \textit{1}    & 15.0(6.05)  & 5.9(2.36)   & & 24.2(17.56)  & 32.9(24.87)  \\
& & \textit{32} & 357.9(2.36) & 228.9(2.64) & & 582.8(20.96) & 793.0(33.89) \\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

\section{Вывод о целесообразности применения того или иного вида разверток для того или иного вида задач}

@@

\bibitem{globalizerSystem}
Gergel V.P., Barkalov K.A., and Sysoyev A.V: Globalizer: A novel supercomputer software system for solving time-consuming global optimization problems. Numerical Algebra, Control \& Optimization 8(1), 47--62, 2018
